Sequence divergence as the main determinant of sequence-structure relationships

### 2015-05-14 manuscript/cordiv/ej_manuscript_cordiv.tex
Making more changes

\begin{abstract}
Recent work has shown that structural properties are capable of predicting site-specific evolutionary rates for a given protein. However, the strength and significance of these structure-sequence relations appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$. Here we present the results from a comprehensive search for potential biophysical and structural determinants of protein evolution by studying more than $200$ structural and evolutionary properties in a dataset of $209$ monomeric enzymes. We discuss the main protein characteristics responsible for the general patterns of protein evolution, and identify sequence divergence as the main determinant of the strengths of virtually all structural-evolution relationships, explaining $10-30 \%$ of observed variation in sequence-structure relations. In addition to sequence divergence, we identify several protein structural properties that are moderately but significantly coupled with the strength of sequence-structure relations. In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to have the strongest sequence-structure relations.

\end{abstract}
\vfill
\vfill
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}

@@

Proteins are subject to a number of biophysical constraints and functional constraints \citep{Scherreretal2012, Wilkeetal2010}. These constraints result site-specific patterns of sequence variability within a protein. Recently several site-specific structural properties that can explain the general patterns of sequence variability in proteins have been identified. One of the earliest examples is Relative Site Accessibility \citep{Ramseyetal2011, Fransozaetal2009}. It was found that residues that are buried in the core of proteins tend to be more conserved than exposed residues close to the surface of the protein.  Another structural property that has been found to significantly correlate with with site-specific variation is local packing density, often measured by Contact Number (CN) or Weighted Contact Number (WCN) \citep{Liaoetal2005, Yehetal2014, Huangetal2014}\\
\indent In a recent work, \cite{Echaveetal2015} presented a biophysical model that links the thermodynamic stability changes due to mutations at each site in proteins (\ddg)  to the rate at which mutations accumulate in the corresponding site over evolutionary time. They found that the variations in the free-energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates, comparable to the predictive powers of other structural variables such as residue solvent accessibility and contact number. \\
\indent Although the majority of proteins exhibit some degree of correlation and association between sequence variation and structural properties, the strength of these correlations vary widely among different structures. By analyzing a data set of $216$ monomeric enzymes, \cite{Yehetal2014} found a wide range of $\rho\sim0.1-0.8$ for the Spearman correlation strengths of sequence variability with two site-specific properties: the Weighted Contact Number (WCN) and RSA. Similarly, \citep{Echaveetal2015} recently found a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg -- with evolutionary rates. It appears that sequence-structure correlations tend to correlate strongly with each other implying that several structural properties are representing similar biophysical constraints on site-specific evolutionary rates(Figure \ref{fig:seqent_structure_cors}). This implies that for a given protein, the correlation strength of a specific structural property with evolutionary rates can serve as a proxy for the correlation strength of other structural properties with sequence evolutionary rates. {\color{red} Where is the stress model, include it!!!}\\
\indent The fact that all relevant structural properties seem to have more or less the same predictive power for sequence variability implies the existence of one or more structural or evolutionary characteristics of proteins that modulate sequence-structure correlations in all proteins. Motivated by these observations, here we present the results of a comprehensive search for the potential underlying structural or evolutionary properties of proteins that can explain the wide range of variations seen in correlation strengths of sequence evolutionary rates with different structural properties.  We show that among all properties considered, sequence divergence appears to be the primary determinant for the strength of sequence-structure relationships. In addition, we show that proteins with more homogeneous Hydrogen bond (H-bond) energies, higher fraction of Helical structures and lower number of $\beta$-sheets generally tend to exhibit the strongest sequence-structure correlations.  \\

@@

    \subsection*{Structures, sequence data, and measures of sequence properties } 
    The results presented in this work are based on a dataset of $209$ monomeric enzymes \citep{Echaveetal2015} randomly picked from the Catalytic Site Atlas $2.2.11$ \citep{Porteretal2004} with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes \citep{Webb1992} and domains of all main SCOP structural classes \citep{Murzinetal1995}. To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by \citep{Yehetal2014, Huangetal2014} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol \citep{Goldenbergetal2008}. Sequence alignments were then constructed using amino-acid sequences with MAFFT \citep{Katohetal2002, Katohetal2005}, specifying the auto flag to select the optimal algorithm for the given data set. The alignments were then used to calculate the site-specific evolutionary rates for each individual protein in dataset. To do so, we relied on two independent methods of measuring sequence variability measure. First, we calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$:
    \begin{equation}
        \label{eqn:shannon}
        H_i = -\sum_j P_{ij}\ln P_{ij}
    \end{equation}

    where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment. The sequence entropy is a measure of variability at each site. We also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software Rate4site {\color{red}(Citation??)}. First the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity cited{Stamatakis2006, Stamatakis2014}. For each structure, we then used the respective sequence alignment and phylogenetic tree to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (aaJC) \citep{Mayroseetal2004}.

    \subsection*{Calculation of Structural Properties}
    The goal of the presented work is to identify the prominent structural or sequence properties of proteins that modulate sequence-structure correlations. These potential modulators represent a unique characteristics of the protein as a whole.  In general, the structural and evolutionary properties fall into two major categories.  1. {\it Residue-level properties}: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence. Prominent examples of the site-specific structural properties include RSA \cite{Fransozaetal2009, Scherreretal2012, Yehetal2014}  and WCN \citep{Shihetal2012, Yehetal2014} {\color{red}Should I cite Lin paper?}.   2. {\it PDB-level properties}: structural or evolutionary characteristics that are representative of the protein as a whole. Examples include protein Contact Order (CO) as defined by {\color{red}Need Citation!}, protein size and compactness, sequence length, and structural resolution of the protein in X-ray crystallography.  In addition, the distribution of each residue-level property can be summarized by its statistical moments as pdb-level property of the protein. Prime examples include, the mean and variance of WCN, RSA, sequence entropy, evolutionary rates. A comprehensive list of protein properties and their definitions are given in Table 
 \ref{tab:definitions} {\color{red} Where is this table???!!!}. \\
\indent A popular tool in condensed matter physics, Voronoi tessellation of a set of points (seeds) is a way of dividing the space into a number of regions such that for each seed there will be a corresponding region consisting of all points closer to that seed than to any other. These regions are called Voronoi cells. The structure of proteins can be considered as a set of 3D coordinates representing individual sites. We use VORO++ software {\color{red}(Need citation!)} to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP \citep{Kabschetal2005}  for the calculation of Accessible Surface Area (ASA) for each site. We normalized the ASA for each site by the theoretical maximum solvent accessibility values of \cite{Tienetal2013}  to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins. In addition to ASA values, we also extract from DSSP output, information about the secondary structure of proteins such as the total number of residues participating in different types of helices, parallel or anti-parallel beta sheets, or loops and turns. To complete the list of pdb-level structural properties, we also calculate the Spearman correlations between all residue-level structure and sequence properties and include them in the analysis to probe their potential effects on the strength of structure-sequence relations. \\

>>>>

\begin{abstract}
Recent work has shown that structural properties are capable of predicting site-specific sequence variability for a given protein. However, the strength and significance of these structure-sequence relations appear to vary widely among different proteins, with absolute correlation strengths ranging from $0.1$ to $0.8$. Here we present the results from a comprehensive search for potential biophysical and structural determinants of protein evolution by studying more than $200$ structural and evolutionary properties in a dataset of $209$ monomeric enzymes. We discuss the main protein characteristics responsible for the general patterns of protein evolution, and identify sequence divergence as the main determinant of the strengths of virtually all structural-evolution relationships, explaining $10-30 \%$ of observed variation in sequence-structure relations. In addition to sequence divergence, we identify several protein structural properties that are moderately but significantly coupled with the strength of sequence-structure relations. In particular, proteins with more homogeneous back-bone hydrogen bond energies, large fractions of helical secondary structures and low fraction of beta sheets tend to have the strongest correlations between structural properties and site variability. 
\end{abstract}
\vfill
\vfill
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}

@@

{\color{blue}
The result in this paper: Site variabilty is the main determine of the difference in structure-sequence correlations. This is the main. It might be good to mention that we looked at 200 other predictors and found this as the more important one. The issue: There is a difference between the Huang et al paper vs. Sharamoradi et al. In Echave's paper with Huang or Yeh?, they say that the CN is the best predictor. In Amir's paper, we say that RSA is still the best predictor. Why is this different? The difference is in the datasets in terms of variability. This paper should have about 3 figures according to Claus possibly. So far I see 4 plots that need to be made: \\
1. WCN-Rate4site vs Mean RateSite??? Does this make sense? (Also for RSA) \\
2. WCN-Entropy vs Mean Entropy \\
3. RSA-Entropy vs Mean Entropy \\
Open Questions: Should I make the stress  and delta delta G correlations plots for the viral sequences that would complete the story?}

Proteins are subject to a number of biophysical and functional constraints \citep{Scherreretal2012, Wilkeetal2010}. These constraints result site-specific patterns of sequence variability within a protein. Recently several site-specific structural properties that can explain patterns of sequence variability in proteins have been identified. One of the earliest examples is Relative Site Accessibility \citep{Ramseyetal2011, Fransozaetal2009}. Residues that are buried in the core of proteins tend to be more conserved than exposed residues close to the surface of the protein.  Another structural property that has been found to significantly correlate with with site-specific variation is local packing density, often measured by Contact Number (CN) or Weighted Contact Number (WCN) \citep{Liaoetal2005, Yehetal2014, Huangetal2014}\\
\indent In a recent work, \cite{Echaveetal2015} presented a biophysical model that links the thermodynamic stability changes due to mutations at each site in proteins (\ddg)  to the rate at which mutations accumulate in the corresponding site over evolutionary time. They found that the variations in the free-energy of the protein due to amino acid substitutions at individual sites can explain the site-specific evolutionary rates comparable to the predictive powers of solvent accessibility and contact number. \\
\indent Although the majority of proteins exhibit some degree of correlation and association between sequence variation and structural properties, the strength of these correlations vary widely among different proteins. By analyzing a data set of $216$ monomeric enzymes, \cite{Yehetal2014} found a wide range of $\rho (\sim0.1-0.8)$ for the Spearman correlation strengths of sequence variability with two site-specific properties: the Weighted Contact Number (WCN) and RSA. Similarly, \cite{Echaveetal2015} recently found a wide range of $\rho\sim0.2-0.8$ for the correlation strength of the site-specific stability contribution -- quantified by \ddg -- with evolutionary rates. {\color{blue} It appears that sequence-structure correlations tend to correlate strongly with each other implying that several structural properties are representing similar biophysical constraints on site-specific evolutionary rates. SHOULD THIS BE IN RESULTS??} (Figure \ref{fig:seqent_structure_cors}). This implies that for a given protein, the correlation strength of a specific structural property with evolutionary rates can serve as a proxy for the correlation strength of other structural properties with sequence evolutionary rates. {\color{red} Where is the stress model, include it!!! Need to mention Huang Paper somewhere!}\\
\indent The fact that all relevant structural properties seem to have more or less the same predictive power for sequence variability implies the existence of one or more structural or evolutionary characteristics of proteins that modulate sequence-structure correlations in all proteins. The underlying principles behind the wide variation in correlation strengths among proteins in not well understood. Here we present the results of a comprehensive search for the potential underlying structural or evolutionary properties of proteins that can explain the wide range of variations seen in correlation strengths of sequence evolutionary rates with different structural properties.  We show that among all properties considered, {\color{blue}sequence divergence. What?? Can we discuss divergence when talking about entropy?} appears to be the primary determinant for the strength of sequence-structure relationships. In addition, we show that proteins with more homogeneous Hydrogen bond (H-bond) energies, higher fraction of Helical structures and lower number of $\beta$-sheets generally tend to exhibit the strongest sequence-structure correlations.  \\

@@

    \subsection*{Structures, sequences, and measures of sequence properties } 
    The results presented in this work are based on a dataset of $209$ monomeric enzymes \citep{Echaveetal2015} randomly picked from the Catalytic Site Atlas $2.2.11$ \citep{Porteretal2004} with protein sizes in the sample ranging from $95$ to $1287$, including representatives from all six main EC functional classes \citep{Webb1992} and domains of all main SCOP structural classes \citep{Murzinetal1995}. To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by \citep{Yehetal2014, Huangetal2014} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol \citep{Goldenbergetal2008}. Sequence alignments were then constructed using amino-acid sequences with MAFFT \citep{Katohetal2002, Katohetal2005}, specifying the auto flag to select the optimal algorithm for the given data set. The alignments were then used to calculate the site-specific evolutionary rates for each individual protein in dataset. To do so, we relied on two independent methods of measuring sequence variability measure. First, we calculated the Shannon entropy ($H_i$) -- the sequence entropy, hereafter abbreviated as {\it seqent} -- at each alignment column $i$:
    \begin{equation}
        \label{eqn:shannon}
        H_i = -\sum_j P_{ij}\ln P_{ij}
    \end{equation}

    where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment. The sequence entropy is a measure of variability at each site. We also calculated a measure of site-specific evolutionary rate -- hereafter abbreviated as {\it r4s} -- for each protein using software Rate4site {\color{red}(Citation??)}. First the Maximum Likelihood phylogenetic trees were inferred with RAxML, using the LG substitution matrix and the CAT model of rate heterogeneity \citep{Stamatakis2006, Stamatakis2014}. For each structure, we then used the respective sequence alignment and phylogenetic tree to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model (aaJC) \citep{Mayroseetal2004}.

    \subsection*{Calculation of Structural Properties}
    The goal of the presented work is to identify the prominent structural or sequence properties of proteins that modulate sequence-structure correlations. These potential modulators represent a unique characteristics of the protein as a whole.  In general, the structural and evolutionary properties fall into two major categories.  1. {\it Residue-level properties}: Site-specific structural or evolutionary properties that are defined and calculated for each specific amino acid site in the protein sequence. Prominent examples of the site-specific structural properties include RSA \citep{Fransozaetal2009, Scherreretal2012, Yehetal2014}  and WCN \citep{Shihetal2012, Yehetal2014} {\color{red}Should I cite Lin paper?}.   2. {\it PDB-level properties}: structural or evolutionary characteristics that are representative of the protein as a whole. Examples include protein Contact Order (CO) as defined by {\color{red}Need Citation!}, protein size and compactness, sequence length, and structural resolution of the protein in X-ray crystallography. {\color{blue} RE-WRITE THIS In addition, the distribution of each residue-level property can be summarized by its statistical moments as protein-level property of the protein}. Prime examples include, the mean and variance of WCN, RSA, sequence entropy, evolutionary rates. A comprehensive list of protein properties and their definitions are given in Table 
 \ref{tab:definitions} {\color{red} Where is this table???!!!}. \\
\indent A popular tool in condensed matter physics, Voronoi tessellation of a set of points (seeds) is a way of dividing the space into a number of regions such that for each seed there will be a corresponding region consisting of all points closer to that seed than to any other. These regions are called Voronoi cells. The structure of proteins can be considered as a set of 3D coordinates representing individual sites. We use VORO++ software {\color{red}(Need citation!)} to calculate the relevant Voronoi cell properties of all sites in all proteins, and use DSSP \citep{Kabschetal2005}  for the calculation of Accessible Surface Area (ASA) for each site. We normalized the ASA for each site by the theoretical maximum solvent accessibility values of \cite{Tienetal2013}  to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins. In addition to ASA values, we also extract from DSSP output, information about the secondary structure of proteins such as the total number of residues participating in different types of helices, parallel or anti-parallel beta sheets, or loops and turns. To complete the list of pdb-level structural properties, we also calculate the Spearman correlations between all residue-level structure and sequence properties and include them in the analysis to probe their potential effects on the strength of structure-sequence relations. \\
